---
params:
  subtitle: "Looking for droughts"
  title: "Weather data"
  authors: "Ben Anderson (b.anderson@soton.ac.uk, `@dataknut`)"
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---


# Set up

```{r setup}

knitr::opts_chunk$set(echo = TRUE)

# Set start time ----
startTime <- proc.time()

# Libraries ----

message("Loading libraries...")

library(here) # where are we?
library(data.table) # data frames with superpowers
library(ggplot2) # pretty plots
library(kableExtra) # pretty tables
library(lubridate) # date & time stuff
library(plotly) # interactive plots
library(skimr) # descriptive & summary stats

# Parameters ----
dPath <- paste0(here::here(), "/data/")
dFile <- paste0(dPath, "ClimateData_1976_2019.csv")

# get data ----

message("Getting data from ", dFile)

weatherDT <- data.table::fread(dFile) # load data
```

# Background

We want to load some UK weather data and have a look at it. We then want to see if we can:

 * find droughts. Droughts in the UK are periods of > 14 days where precipitation is lower than 0.2 (cm?);
 * ....

# Have a look at the data

```{r checkData}
weatherDT$V1 <- NULL # what's the row number for?

# do some pre-processing ----
weatherDT[, rDate := lubridate::dmy(Date)] # make nice date
weatherDT[, year := lubridate::year(rDate)] # make nice year
weatherDT[, month := lubridate::month(rDate, label = TRUE, abbr = TRUE)] # make nice date
t <- head(weatherDT) # what have we got?

kableExtra::kable(t, caption = "First few rows of processed data") %>%
  kable_styling()

skimr::skim(weatherDT)
```


Just for fun let's plot the data. You should be able to hover the mouse over the dots in \@ref(fig:plotData) to investigate.

```{r plotData, fig.cap="Mean precipitation per month by year"}
# just for fun
plotDT <- weatherDT[, .(meanPrcpt = mean(prcpt)), 
                    keyby = .(year, month)]

p <- ggplot2::ggplot(plotDT, aes(x = month, y = meanPrcpt, group = year, colour = year)) +
  geom_point() +
  scale_color_viridis_c() +
  labs(y = "Mean precipitation",
       caption = "Year treated as continuous for pretty colours")

plotly::ggplotly(p) # interactive plot
```

So that looks vaguely right.

# Challenge 1: Find the droughts

Count the number of spans (per year?) of 15 or more days when the daily rainfall < 0.2. This is the UK definition of drought.

To do this we:

 * mark the first & last days in a sequence of low rainfall (< 0.2) days
 * get the time difference between them in days
 * find the ones where the difference is > 14

Table \@ref(tab:findDroughts) shows the number of low precipitation events (1 or more days with < 0.2) which were droughts (> 14 days in a row) by start month and year. Perhaps surprisingly 1976 has none.

```{r findDroughts}

setkey(weatherDT,rDate) # ensures ordered by date
weatherDT[, droughtThresh := ifelse(prcpt < 0.2, 1, 0)] # to make life easier
# flag the start & ends of low precipitation periods
weatherDT[, droughtPeriod := ifelse(droughtThresh == 1 & 
                                      shift(droughtThresh == 0), # first of 1 -> n days with low
                                   "Start", 
                                   NA)] 
weatherDT[, droughtPeriod := ifelse(droughtThresh == 0 & 
                                      shift(droughtThresh == 1), # last of 1 -> n days with low
                                    "End", 
                                    droughtPeriod)] 

# now isolate the start & ends of low precipitation periods
lowPrcptPeriodsDT <- weatherDT[droughtPeriod == "Start" | 
                                 droughtPeriod == "End"]

lowPrcptPeriodsDT[, periodCount := ifelse(droughtPeriod == "Start",
                                          shift(rDate, type = "lead") - rDate, # n days between start & end (they are ordered)
                                          NA) # undefined between end & start (we don't care about periods between droughts for now)
                  ] 

# flag the droughts using the 14 day threshold
lowPrcptPeriodsDT[, drought := ifelse(periodCount > 14, 
                                      "Drought start", 
                                      "Not drought")]

plotDT <- lowPrcptPeriodsDT[, .(nDroughts = .N), # data.table magic
                            keyby = .(drought, year, start = month)] # add up number of droughts/no droughts in these periods of low precip

# save out the low precip periods
outF <- paste0(dPath, "lowPrecipSequences.csv")
data.table::fwrite(lowPrcptPeriodsDT, outF) # includes start and end of sequences
# save out just the flagged droughts
lowPrcptPeriodsDT[, drought := ifelse(droughtPeriod == "End" &
                                        shift(periodCount > 14),
                                      "Drought end",
                                      drought)
                  ] # flag drought end as drought too for data output
outF <- paste0(dPath, "lowPrecipSequencesDroughts.csv")
data.table::fwrite(lowPrcptPeriodsDT[drought %like% "Drought"], outF)

ggplot2::ggplot(lowPrcptPeriodsDT, aes(x = periodCount)) +
  geom_histogram() +
  labs(x = "Duration of low precipitation periods (sequences of days < 0.2 cm, entire period)")
```

```{r droughtsTable}
kableExtra::kable(plotDT[drought == "Drought start"], caption = "Number of low precipitation periods that were droughts by year") %>%
  kable_styling()
```

Just to make this clear, Figure \@ref(fig:plotDroughts) shows the pattern of number of droughts by start month and year.

```{r plotDroughtCounts, fig.cap="Number of droughts by start month and year"}
p <- ggplot2::ggplot(plotDT[drought == "Drought start"], 
                aes(x = start, fill = nDroughts, y = year)) +
  geom_tile() +
  labs(caption = "Note: months without droughts (Jan/Feb & Dec) not shown")

p
```

If this chart in not what you expect then remember:

 * the plot is **not** showing the actual span of days, it is showing a count of how many drought events there were that started in a given month;
 * the method will not classify a sequence of 4 days 'rain < 0.2 cm' -> 1 day 'rain > 0.2' -> 11 days 'rain < 0.2' as drought. It is, literally just finding periods of > 14 consecurive days with 'rain < 0.2'. It seems there weren't any in 1976 despite it being a "drought" (see below);
 * meteorological drought (no rain) and hydrological drought (no ground water) are not the same thing...
 * our experience of drought and what plants experience as water stress may not correlate that closely with the meteorological definition of drought used here...
 
 Of note:

 * 1976: Was a hot summer after a dry winter (1975-1976) - so this was a hydrological drought - there was rain, but not enough to refill the reservoirs... https://historicdroughts.ceh.ac.uk/content/northern-drought-1984
 * 1984: https://historicdroughts.ceh.ac.uk/content/northern-drought-1984 & https://historicdroughts.ceh.ac.uk/content/northern-drought-1984
 * 1995: https://en.wikipedia.org/wiki/1995_Great_Britain_and_Ireland_heat_wave & https://historicdroughts.ceh.ac.uk/content/tanker-drought-1995-1998
 * 2003: https://en.wikipedia.org/wiki/2003_European_heat_wave#United_Kingdom & https://historicdroughts.ceh.ac.uk/content/hot-summer-2003 

We can also plot the drought spans themselves (which might have been more useful in any case :-) - see Figure \@ref(fig:plotDroughtPeriodsSegment).

```{r plotDroughtPeriodsHisto, fig.cap="Drought periods histogram"}

droughtsDT <- lowPrcptPeriodsDT[drought %like% "Drought"]

ggplot2::ggplot(droughtsDT, aes(x = periodCount)) +
  geom_histogram() +
  labs(x = "Drought duration (days) - whole period")

t <- droughtsDT[, .(meanDuration = mean(periodCount, na.rm = TRUE),
                    minDuration = min(periodCount, na.rm = TRUE),
                    maxDuration = max(periodCount, na.rm = TRUE))]

kableExtra::kable(t, caption = "Duration stats (all years)") %>%
  kable_styling()
```

```{r plotDroughtPeriodsSegment, fig.cap="Drought periods over time"}
droughtsDT[, startDate := rDate] 
droughtsDT[, endDate := shift(rDate, type = "lead")]

# so how do we plot that?
plotDT <- droughtsDT[drought == "Drought start"]
plotDT[, startDec := lubridate::date_decimal(2000 + (lubridate::decimal_date(startDate) - year))]
plotDT[, endDec := lubridate::date_decimal(2000 + (lubridate::decimal_date(endDate) - year))]

p <- ggplot2::ggplot(plotDT, aes(x = startDec, xend = endDec,
                                y = year, yend = year)) +
  geom_segment(colour = "red") +
  # theme(legend.position="bottom") +
  # guides(colour = guide_legend(title = "Year", nrow = 2)) +
  scale_y_continuous(breaks = c(1975, 1980, 1985,1990,1995,2000,2005,2010,2015,2020)) +
  labs(x = "Date") +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%b %d")
p +
  geom_text(
    label= plotDT$year, 
    size = 3,
    nudge_y = 0.5,
    check_overlap = T
  )
```


# Runtime


```{r check runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes) using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform`.

# R environment

R packages used:

 * base R - for the basics [@baseR]
 * data.table - for fast (big) data handling [@data.table]
 * ggplot2 - for slick graphics [@ggplot2]
 * here - for here [@here]
 * lubridate - date manipulation [@lubridate]
 * kableExtra - fancy tables [@kableExtra]
 * knitr - to create this document [@knitr]
 * skimr - data summaries [@skimr]
 
```{r sessionInfo}
sessionInfo()
```


# References


